Use and acquisition of causal knowledge in a decision making process

### 2018-11-24 Propuesta/Presentacion/Presentacion_Corta.tex
presentacion version sabado

}
\usepackage{booktabs}
\usepackage[utf8]{inputenc}
\usepackage[spanish, mexico]{babel}
\usepackage{listings}
\usepackage{breakcites}
\usepackage{dsfont}
\usepackage{hyperref}
\usepackage{amssymb,amsthm,amsmath,latexsym}
\usepackage{natbib}
\theoremstyle{plain}
\newtheorem{teo}{Teorema}
\newtheorem{prop}[teo]{Proposición}
\newtheorem{defi}[teo]{Definición}
\newtheorem{obs}[teo]{Observación}
\newtheorem{lem}[teo]{Lema}
\newtheorem{cor}[teo]{Corolario}
\usepackage{tikz}
\usetikzlibrary{trees}
\usetikzlibrary{calc}

@@

\section{Introduction}
\subsection{General Overview}
\begin{frame}
\frametitle{General Idea}
\begin{itemize}
\item We live in a causally structured world. 
\item Causal relations allow us to plan and to reason.
\item Causal reasoning allows for \textit{what...if...} claims.
\item In order to engage in causal reasoning we must be able to \textit{imagine} alternative worlds. 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Engaging in a Causal Environment}
\begin{itemize}
\item If Causal relations allow for all of this, why don't we use them?
\item In fact, we do, although not necessarily we are aware of this. 
\item \cite{hagmayer2013repeated} show how human beings use and modify causal information.
\item \cite{hohwy2013predictive}, \cite{clark2015surfing}, asserts that the brain itself is a causal hypothesis tester. 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Our Problem}
\begin{itemize}
\item Engaging with the world has many dimensions: perception, action, agency, etc.
\item We consider only the decision making process where consequences are uncertain.
\item Consequences and actions are causally related.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Our setting}
\begin{itemize}
\item Consider a \textit{rational} decision maker in an uncertain environment.
\item This is, consequences can not be predicted, which is equivalent to saying that making the same choice does not has the same consequence.
\item Examples: stock picking.
\item We add an extra to the environment: causal structure. 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Our Proposal}
\begin{itemize}
\item Since decision-making theories do not consider causal structure in the decision making processes
\item We \textbf{propose} an autonomous rational agent who can use and learn causal information while interacting with its environment.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Rationality and Causality}
\begin{itemize}
\item Rationality is defined in terms of \textit{axioms of coherence}.
\item From the axioms, a formal criteria of choice is obtained: \textbf{maximization of expected utility}. 
\item How is the awareness of causal structure and the rationality assumption related?
\item \cite{joyce1999foundations} stated that a decision maker should consider effects of of decisions, although he didn't said how. 
\item We want to answer how to maximize expected utility in terms of a causal model. 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Our idea}
\begin{itemize}
\item A decision maker who is aware of the causal nature of his environment must use those relations to predict outcomes and take the actions that renders a good (prefered) consequence. 
\item The agent will use current knowledge in order to take best possible action and then update causal \textit{beliefs} after observing outcome.
\item Learning causal information by updating beliefs while interacting with environment.
\end{itemize}
\end{frame}

\subsection{Outline}
\begin{frame}
\frametitle{Outline}
\begin{itemize}
\item We now have two main issues to talk about: Causality and Decision Making. 
\item Once we have set the language, then we will state the problem we are trying to attack.
\item Remember, if you have an idea, then probably someone else already had it: related work.
\item Our attempt, and some results that we have stumbled upon. 
\end{itemize}
\end{frame}

\section{Framework and Main Concepts}
	\subsection{Causality}
	\begin{frame}
		\frametitle{What is causality?}
		\begin{itemize}
		\item We want to think about the causes of certain events; i.e., a disease, an accident. 
		\item Causal claims are associated with what didn't happen. 
		\end{itemize}
	\end{frame}
		\begin{frame}
		\frametitle{Nature of Causality}
		\begin{itemize}
		\item Do causal relations \textit{really} exist? 
		\item Or are they only a mental construct useful for managing certain regularities in our world?
		\item We prefer the former idea. 
		\end{itemize}
		\end{frame}

		\begin{frame}
		\frametitle{Difference between causal and associative relations}
		\begin{itemize}
		\item With associative relations we can't figure out what influenced (or caused) what.
		\item Consider this: A distribution $p(a,t)$ can be factored both as $p(a|t)p(t)$ and $p(t|a)p(a)$. 
		\item In terms of a Bayesian Network, $T \to A$, and $A \to T$, but both can't be causal.
		\item Causal relations need not be in temporal order: consider a barometer and a storm.
		\end{itemize}
		\end{frame}

		\begin{frame}
		\frametitle{The Ladder of Causality (\cite{pearlwhy})}
		Pearl, in his recent book \cite{pearl2018why} considers three levels of scientific inquiry.
		\begin{enumerate}
		\item Observing
		\item Intervening
		\item Counterfactual reasoning
		\end{enumerate}
		\end{frame}

		\begin{frame}
		\frametitle{Definition of Causality}
		The working definition of Causality that we will be using is based upon \cite{spirtes2000causation} and \cite{pearl2009causality}.
		\end{frame}

		\begin{frame}
		\frametitle{Formal Definition}
		\begin{defi}
		Let $(\Omega, \mathcal{F}, \mathbb{P})$ a finite, probability space and consider a binary relation  $\to  \subseteq \mathcal{F} \times \mathcal{F}$ such that
		\begin{itemize}
		\item Transitive: if $A \to B$ and $B \to C$ for $A,B,C \in \mathcal{F}$ then $A \to C$.
		\item  Irefflexive: for each $A \in \mathcal{F}$ it is not true that $A \to A$.
		\item Antisymetric: For $A, B \in \mathcal{F}$, if $A \to B$, then it is not true that $B \to A$.
		\end{itemize}
		If $A \to B$ we will say that $A$ is a cause of $B$.
		\end{defi}
		\end{frame}
		\subsubsection{Causal Graph Construction}
		\begin{frame}
		\frametitle{From Causes to Graphs}
		For each pair $A,B \in \mathcal{F}$ such that $A \to B$ and $A$ is a direct cause of $B$, define a node representing each variable, and a directed edge joining them.
		\begin{prop}
		Given a binary relation “$\to$” that satisfies previous requirement, let $\mathcal{G}$ be the directed graph that is obtained by the previous procedure. Then, $\mathcal{G}$ contains no cycles.
		\end{prop}
		\begin{proof}
		Consider a cycle $A \to C_1 \to C_2 \to \cdots \to A$, then we would have that $A \to A$, which contradicts the Irreflexive property.
		\end{proof}
		\end{frame}

		\begin{frame}
		\frametitle{Endogenous and Exogenous variables}
		\begin{obs}
		Notice that if the graph is finite, then necesarily there will be nodes without parents, which in the context of causal models will be called exogenous variables.
		\end{obs}
		\end{frame}

		\begin{frame}
		\frametitle{Causal Sufficiency}
		A requirement needed for the following construction, known as Causal Sufficiency, says that any cause of any event in the model is in the model. 
		\end{frame}

		\subsubsection{Probability Measure built from the Graph}
		\begin{frame}
		\frametitle{from Graphs to Probability Measures}
		Given a Directed Acyclic Graph, one can build the probability distribution $P_\mathcal{G}$ that expresses the relations of the graph. We require that this measure satisfies:
        \begin{itemize}
        \item Causal Markov Condition.
        \item Causal Minimality.
         \item Causal Faithfulness.
        \end{itemize}
         This requirements will be treated in an axiomatic fashion.
		\end{frame}

	\subsection{Causal Graphical Models}
	\begin{frame}
	\frametitle{Causal Graphical Models}
	\begin{defi}
	A causal graphical model consists of:
	\begin{itemize}
	 \item A set of random variables $\mathcal{X}=\{ X_1,...,X_n \}$.
	 \item A directed acyclic graph (DAG) $\mathcal{G}$.
	 \item An operator $do()$ defined over the space of DAG's whose action consists of: given $\mathbf{X} \subseteq \mathcal{X}$ and $\mathbf{x} = \{ x_{i_1}, x_{i_2}, ... , x_{i_j} \} \in Val(\mathcal{X})$, then $do(\mathbf{X} = \mathbf{x} )$ assigns to each $X_j \in \mathbf{X}$ the value $x_{i_j}$ and deletes any incoming edge on it.
	 \end{itemize}
	 \end{defi}
	 \end{frame}

	 \begin{frame}
	 \frametitle{Causation in terms of the Do operator}
	 \begin{itemize}
	 \item We can use the Do operator to provide a definition of a cause:
	 \begin{defi}
	 We say that $X$ \textit{causes} $Y$ if 
	\[ P_\mathcal{G}(Y | do(X=x)) \neq P_\mathcal{G}(Y | do(X=x')). \]
	 \end{defi}
	 \end{itemize}
	 \end{frame}

	\subsection{Decision Theory}
	\begin{frame}
	\frametitle{Decision Problems}
	A decision problem is composed by:
	\begin{itemize}
	\item A decision maker.
	\item A set of available options $\mathcal{A}$.
	\item A preference relation $\succeq$ defined over $\mathcal{A} \times \mathcal{A}$.
	\item A set of uncertain events $\mathcal{E}$.
    \item A set of possible outcomes $\mathcal{C}$
	\end{itemize}
	\end{frame}

	\begin{frame}
	\frametitle{Rationality}
	Rationality is a subtle concept with intrinsec philosophical implications. We will understand by \textit{rational} a decision maker whose \textit{preferences} are some how logically consistent.
	In its most basic form:
	\begin{itemize}
	\item The decision maker needs to be able to strictly prefer at least between one action over other.
	\item Transitive preferences.
	\item If $a$ is always prefered over $b$, then it doesn't matter what uncertain event occurs.
	\end{itemize}
	\end{frame}

	\begin{frame}
	\frametitle{Expected Utility Maximizaton.}
	We are considering \textit{bounded decision problems};i.e., where a \textit{best} and \textit{worst} outcomes exist. In that case,
	\begin{teo}
	For any bounded decision problem with extreme consequences $c^\ast \succ c_\ast$ and any $G \succ \emptyset$
	\[ a_1 \succeq_G a_2 \Leftrightarrow \bar{u}(a_1) \geq \bar{u}(a_2). \]
	\end{teo}
	\begin{proof}
	See \cite{bernardo2000bayesian} Chapter 2 “Foundations”, Proposition 2.22.
	\end{proof}
	\end{frame}

\section{Research Problem}
\begin{frame}
\frametitle{Section:Research Problem}
In this section we use the mathematical formalism described above in order to formally enounce the problem we are trying to solve here and the questions that motivated that problem. 
\end{frame}

	\subsection{Research Questions}
	\begin{frame}
	\frametitle{Research Questions}
	The proposed research is ultimately trying to answer a series of questions about the nature of causal relations and about their possible use.
	\begin{enumerate}
	\item How a rational decision maker who faces an uncertain environment which is governed by a causal mechanism can learn and make use of this causal structure in order to make good    	choices? 
	\item How can a causal structure help a decision maker in order to guide his learning process? 
	\item What does the rationality assumption implies about how to choose when considering causal information? 
\item How to trade off exploration and exploitation when trying to learn about the causal structure of an environment while also trying to make good choices?
\end{enumerate}
\end{frame}
	\subsection{Statement}
	\begin{frame}
	\frametitle{Problem Statement}
	\begin{itemize}
	\item Let  $\mathcal{G}$ a causal graphical model.
	\item Let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem under uncertainty whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$. 
	\item Consider a rational decision maker who doesn't know the causal model which control the probabilities of observing a consequence given an action $a \in \mathcal{A}$.
	\item We ask how the decision maker could learn about the causal structure that controls his environment in order to make good choices with respect to the decision makers' preferences.
	\end{itemize}
	\end{frame}
	\subsection{Motivation}
	\begin{frame}
	\frametitle{Motivation}
	\begin{itemize}
	\item Decision making under uncertain conditions is a fundamental part of intelligent reasoning (\cite{lake2017building}).
	\item Intelligent agents often face situations where an action must be chosen in the presence of uncertain conditions.
	\item It is known that human beings conceive their actions on the world as \textit{intervening} in the world (\cite{hagmayer2009decision}).
	\item Following this idea, \cite{lattimoreNIPS2016} consider decision problems where the action to be chosen is an intervention over a known causal graphical model.
	\item What about not knowing the causal model? 
	\end{itemize}
	\end{frame}
	\subsection{Justification}
	\begin{frame}
	\frametitle{Justification}
	\begin{itemize}
	\item Many real-world applications of decision making are solved by \textit{associative} methods which capture only statistical patterns that are found in data.
	\item Current methods in Reinforcement Learning, and specifically in Deep Reinforcement Learning, although good performance in the task, they can not explain \textit{why} a specific trajectory was chosen by the algorithm.
	\item This is highly relevant in real-world applications such as self-driving cars.
	\item Methods based in Deep Neural Networks aren't supposed to explain why a certain output was produced since those methods are based in \textit{parallel distributed representations}.
	\end{itemize}
	\end{frame}
	\begin{frame}
	\begin{itemize}
	\item On the other hand, learning a causal model of an environment and using it to act upon the environment allows to \textit{explain} aspects of the model that a purely associative model would not be abel to explain. 
	\item It allows to ask \textit{why}.
		\end{itemize}
	\end{frame}
	\subsection{Hypothesis}
	\begin{frame}
	\frametitle{Hypothesis}
	\begin{itemize}
	\item Let $\mathcal{G}$ a causal graphical model and let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$.
	\item Then, if a decision maker doesn't know the probabilities of the uncertain events in $E$, by repeatedly making decisions he can learn a causal model of the environment and use it to find the optimal action (in the sense of expected utility theory) in less, or equal, rounds than if he doesn't consider causal information.
	\end{itemize}
	\end{frame}
	\subsection{Objectives}
	\begin{frame}
	\frametitle{General Objectives}
	The proposed research has as a general objective:
	\begin{itemize}
 \item to understand what are the implications of causality for rational decision maker who faces an uncertain environment and how causal relations can be discovered and used in order to make good choices that maximize the expected utility for the decision maker.
	\end{itemize}
	\end{frame}

\section{Approaches and Related Work}
\subsection{Approaches and Related Work: Reinforcement Learinng}
\begin{frame}
\frametitle{Reinforcement Learning (RL)}
\begin{itemize}
\item Reinforcement Learning (RL) has been the standard setting for decision-making and learning by interaction.
\item  In this setting it is assumed that a Markov Decision Process (MDP) is able to capture the effects of the agent's actions in the environment.
\item Reinforcement Learning uses only associative patterns for finding an optimal policy and this methods can not learn beyond what is expressable by an MDP, which is constrained into associative rules and can not express any other kind of structure, such as causal relations. 
\item it can be shown that optimal policies actually achieve the maximum expected utility as required by the rationality condition (\cite{webb2007game}).
\item In this way, when only associative information is available, RL is a coherent learning framework. On the other hand, it remains to be answered how to incorporate a causal model into an on-line decision making algorithm.
\end{itemize}
\end{frame}
\subsection{Approaches and Related Work: Causal learning in humans}
\begin{frame}
\frametitle{Causal Learning in Human beings}
\begin{itemize}
\item Human beings, while facing a sequential decision problem, use and modify causal information.
\item It is also known that human beings focus on \textit{local} aspects while learning causal relations that are later unified into a single structure.
\item  Following this idea, \cite{wellen2012learning} propose a model to explain how observations  and interventions are used by human beings to learn causal structure when little prior information is available. 
\end{itemize}
\end{frame}
\subsection{Approaches and Related Work: Joyce's Causal Decision Theory}
\begin{frame}
\frametitle{Joyce's Causal Decision Theory}
\begin{itemize}
\item An attempt to formalize Decision Theory in the presence of Causal Information was attempted by Joyce, who stated that a decision maker must choose whatever action is more likely to (causally) produce desired outcomes.
\item  His formulation falls short since he does not consider any kind of formal causal model beyond what is commonly understood by causality
\item Although he captured the intuition that causal relations may be used to control the environment and to predict what is caused by the actions of a decision maker.
\end{itemize}
\end{frame}
\subsection{Approaches and Related Work: Bandits}
\begin{frame}
\frametitle{Bandits}
\begin{itemize}
\item In the class of decision problems considered for this proposal it is assumed that the a rational agent must choose some available action and then receive a reward from it.
\item Problems of this nature have been modelled as \textit{bandit} problems.
\item A bandit is an analogy of a \textit{slot machine}.
\item Best-arm algorithms exist: none consider causal information.
\end{itemize}
\end{frame}
\begin{frame}
\frametitle{Causal bandits}
\begin{itemize}
\item n action could be conceived as an \textit{intervention} over some environment, which is in fact how humans consider actions in the world
\item they use and modify causal knowledge during a sequential decision making process.
\item It is shown by \cite{lattimoreNIPS2016} that adding causal information in a fixed budget decision problem allows the decision maker to learn \textit{faster} had he not considered causal information.
\item Their work requires that the causal model is fully known to the decision maker.
\item this requirement is relaxed later by \cite{sen2017identifying} who requires only that some part of the causal model is known and allow interventions over the unknown part. 
\end{itemize}
\end{frame}

>>>>

}
\usepackage{booktabs}
\usepackage[utf8]{inputenc}
%\usepackage[spanish, mexico]{babel}
\usepackage{listings}
\usepackage{breakcites}
\usepackage{dsfont}
\usepackage{hyperref}
\usepackage{amssymb,amsthm,amsmath,latexsym}
\usepackage{natbib}
\theoremstyle{plain}
\newtheorem{teo}{Teorema}
\newtheorem{prop}[teo]{Proposition}
\newtheorem{defi}[teo]{Definition}
\newtheorem{obs}[teo]{Observation}
\newtheorem{lem}[teo]{Lemma}
\newtheorem{cor}[teo]{Corolary}
\usepackage{tikz}
\usetikzlibrary{trees}
\usetikzlibrary{calc}

@@

\section{Introduction}
\subsection{General Overview}
\begin{frame}
\frametitle{General Idea}
\begin{itemize}
\item We live in a causally structured world. 
\item Our actions \textit{cause} changes in the environment.
\item We consider this changes in order to plan our actions
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Human use of Causal Relations}
\begin{itemize}
\item \cite{hagmayer2013repeated} show how human beings use and modify causal information.
\item Both \cite{hohwy2013predictive}, \cite{clark2015surfing}, assert that the brain itself is a causal hypothesis tester. 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Our setting}
\begin{itemize}
\item Consider a \textit{rational} decision maker in an uncertain environment.
\item Consequences of actions can not be predicted.
\item We add an extra to the environment: causal structure. 
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{A brief on causal models}
\begin{itemize}
\item Causal models encode cause-effect relations. 
\item Relations are encoded in a graph.
\item Can be used to predict \textit{effects} of certain causes
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Causal Model: Example}
\begin{figure}[ht]
\vskip 0.2in
\begin{center}
\includegraphics[width=\linewidth]{/Users/MauricioGS1/INAOE/Propuesta/presentacion/example.png}
\caption{Causal graphical model for the test scenario: the target variable \textit{Lives} is causally influenced by the disease the patient has, the treatment assigned and the survival to the secondary effects of treatment.}
\label{causal_model}
\end{center}
\vskip -0.2in
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Our Proposal}
\begin{itemize}
\item Current decision-making theories do not consider causal structure in the decision making processes
\item We \textbf{propose} an autonomous rational agent who can use and learn causal information while interacting with its environment.
\end{itemize}
\end{frame}

\section{Concepts}
\subsection{Causal Models}
\begin{frame}
\frametitle{Concepts}
Concepts and Framework
\end{frame}
		\begin{frame}
		\frametitle{Difference between causal and associative relations}
		\begin{itemize}
		\item With associative relations we can't figure out what influenced (or caused) what.
		\item Consider this: A distribution $p(a,t)$ can be factored both as $p(a|t)p(t)$ and $p(t|a)p(a)$. 
		\item In terms of a Bayesian Network, $T \to A$, and $A \to T$, but both can't be causal.
		\item Causal relations need not be in temporal order: consider a barometer and a storm.
		\end{itemize}
		\end{frame}

		\begin{frame}
		\frametitle{Formal Definition of Causality according to \cite{spirtes2000causation}}
		\begin{defi}
		Let $(\Omega, \mathcal{F}, \mathbb{P})$ a finite, probability space and consider a binary relation  $\to  \subseteq \mathcal{F} \times \mathcal{F}$ such that
		\begin{itemize}
		\item Transitive: if $A \to B$ and $B \to C$ for $A,B,C \in \mathcal{F}$ then $A \to C$.
		\item  Irefflexive: for each $A \in \mathcal{F}$ it is not true that $A \to A$.
		\item Antisymetric: For $A, B \in \mathcal{F}$, if $A \to B$, then it is not true that $B \to A$.
		\end{itemize}
		If $A \to B$ we will say that $A$ is a cause of $B$.
		\end{defi}
		\end{frame}

		\begin{frame}
		\frametitle{From Graphs to Probability Measures}
		Given a Directed Acyclic Graph, one can build the probability distribution $P_\mathcal{G}$ that expresses the relations of the graph. We require that this measure satisfies:
        \begin{itemize}
        \item Causal Markov Condition.
        \item Causal Minimality.
         \item Causal Faithfulness.
         \item Causal Sufficiency
        \end{itemize}
         This requirements will be treated in an axiomatic fashion.
		\end{frame}

	\begin{frame}
	\frametitle{Causal Graphical Models}
	\begin{defi}
	A causal graphical model (\cite{spirtes2000causation}, \cite{pearl2009causality}, \cite{koller2003multi}, \cite{sucar2015probabilistic}) consists of:
	\begin{itemize}
	 \item A set of random variables $\mathcal{X}=\{ X_1,...,X_n \}$.
	 \item A directed acyclic graph (DAG) $\mathcal{G}$.
	 \item An operator $do()$ defined over the space of DAG's whose action consists of: given $\mathbf{X} \subseteq \mathcal{X}$ and $\mathbf{x} = \{ x_{i_1}, x_{i_2}, ... , x_{i_j} \} \in Val(\mathcal{X})$, then $do(\mathbf{X} = \mathbf{x} )$ assigns to each $X_j \in \mathbf{X}$ the value $x_{i_j}$ and deletes any incoming edge on it.
	 \end{itemize}
	 \end{defi}
	 \end{frame}

\subsection{Decision Problems}
	\begin{frame}
	\frametitle{Decision Problems}
	A decision problem is composed by (\cite{bernardo2000bayesian}):
	\begin{itemize}
	\item A (rational) decision maker.
	\item A set of available options $\mathcal{A}$.
	\item A preference relation $\succeq$ defined over $\mathcal{A} \times \mathcal{A}$.
	\item A set of uncertain events $\mathcal{E}$.
    \item A set of possible outcomes $\mathcal{C}$
	\end{itemize}
	\end{frame}

	\begin{frame}
	\frametitle{Choosing rationally}
	Savage's Theorem (\cite{savage1954the}) says that:
\begin{teo}	
	If a rational decision maker does not know the probabilities of uncertain events ocurring, then he must choose according to the expected utility with respect to a \textit{subjective} probability measure and a \textit{subjective} utility function.
\end{teo}
Details can be found at \cite{bernardo2000bayesian}, \cite{gilboa2009decision}.
	\end{frame}

\section{Research Problem}
\begin{frame}
About my Proposed Research
\end{frame}
	\subsection{Motivation}
	\begin{frame}
	\frametitle{Motivation: 1}
	\begin{itemize}
	\item Decision making under uncertain conditions is a fundamental part of intelligent reasoning (\cite{lake2017building}).
	\item Intelligent agents often face situations where an action must be chosen in the presence of uncertain conditions.
	\end{itemize}
	\end{frame}
	\begin{frame}
	\frametitle{Motivation: 2}
	\begin{itemize}
	\item It is known that human beings conceive their actions on the world as \textit{intervening} in the world (\cite{hagmayer2009decision}).
	\item Following this idea, \cite{lattimoreNIPS2016} consider decision problems where the action to be chosen is an intervention over a known causal graphical model.
	\item What about not knowing the causal model? 
	\end{itemize}
	\end{frame}

	\subsection{Justification}
	\begin{frame}
	\frametitle{Justification: Un-interpretable models}
	\begin{itemize}
	\item Many real-world applications of decision making are solved by \textit{associative} methods.
	\item Associative methods capture only statistical patterns that are found in data.
	\item Methods based in Deep Neural Networks aren't supposed to explain why a certain output was produced since those methods are based in \textit{parallel distributed representations}.
	\end{itemize}
	\end{frame}

	\begin{frame}
	\frametitle{Justification: Need for nterpretable decision-making models}
	\begin{itemize}
	\item Current methods for Decision Making are based in Reinforcement Learning, and specifically in Deep Reinforcement Learning. 
	\item Although good performance in the task, they can not explain \textit{why} a specific trajectory was chosen by the algorithm.
	\item This is highly relevant in real-world applications such as self-driving cars.
	\end{itemize}
	\end{frame}

	\begin{frame}
	\frametitle{Justification: Explaining why}
	\begin{itemize}
	\item On the other hand, learning a causal model of an environment and using it to act upon the environment allows to \textit{explain} aspects of the model that a purely associative model would not be abel to explain. 
	\item It allows to ask \textit{why}.
	\end{itemize}
	\end{frame}

	\subsection{Research Questions}
	\begin{frame}
	\frametitle{Research Questions}
	The proposed research is ultimately trying to answer a series of questions about the nature of causal relations and about their possible use.
	\begin{enumerate}
	\item How a rational decision maker who faces an uncertain environment which is governed by a causal mechanism can learn and make use of this causal structure in order to make good    	choices? 
	\item How can a causal structure help a decision maker in order to guide his learning process? 
	\item What does the rationality assumption implies about how to choose when considering causal information? 
\item How to trade off exploration and exploitation when trying to learn about the causal structure of an environment while also trying to make good choices?
\end{enumerate}
\end{frame}

\subsection{Statement}
	\begin{frame}
	\frametitle{Problem Statement}
	\begin{itemize}
	\item Let  $\mathcal{G}$ a causal graphical model.
	\item Let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem under uncertainty whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$. 
	\item Consider a rational decision maker who doesn't know the causal model which control the probabilities of observing a consequence given an action $a \in \mathcal{A}$.
	\item We ask how the decision maker could learn about the causal structure that controls his environment in order to make good choices with respect to the decision makers' preferences.
	\end{itemize}
	\end{frame}

	\subsection{Hypothesis}
	\begin{frame}
	\frametitle{Hypothesis}
	\begin{itemize}
	\item Let $\mathcal{G}$ a causal graphical model and let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$.
	\item Then, if a decision maker doesn't know the probabilities of the uncertain events in $E$, by repeatedly making decisions he can learn a causal model of the environment and use it to find the optimal action (in the sense of expected utility theory) in less, or equal, rounds than if he doesn't consider causal information.
	\end{itemize}
	\end{frame}

\section{Approaches and Related Work}
\begin{frame}
\frametitle{Approaches and Related Work}
\begin{figure}[ht]
\vskip 0.2in
\begin{center}
\includegraphics[width=\linewidth]{/Users/MauricioGS1/INAOE/Propuesta/presentacion/Related_Work.png}
\caption{Causal graphical model for the test scenario: the target variable \textit{Lives} is causally influenced by the disease the patient has, the treatment assigned and the survival to the secondary effects of treatment.}
\label{causal_model}
\end{center}
\vskip -0.2in
\end{figure}
\end{frame} 
\subsection{Approaches and Related Work: Reinforcement Learning}
\begin{frame}
\frametitle{Reinforcement Learning (RL)}
\begin{itemize}
\item Reinforcement Learning uses only associative patterns for finding an optimal policies.
\item Optimal policies actually achieve the maximum expected utility as required by the rationality condition (\cite{webb2007game}).
\item In this way, when only associative information is available, RL is a coherent learning framework. 
\end{itemize}
\end{frame}

\subsection{Approaches and Related Work: Causal learning in humans}
\begin{frame}
\frametitle{Causal Learning in Human beings}
\begin{itemize}
\item Human beings, while facing a sequential decision problem, use and modify causal information.
\item \cite{hagmayer2009decision}, \cite{hagmayer2013repeated} show that human beings can learn causal information in simple decision-making tasks, but remains open to study causal learning while seeking for a specific objective. 
\item 
\end{itemize}
\end{frame}
\subsection{Approaches and Related Work: Joyce's Causal Decision Theory}
\begin{frame}
\frametitle{Joyce's Causal Decision Theory}
\begin{itemize}
\item An attempt to formalize Decision Theory in the presence of Causal Information was attempted by Joyce, who stated that a decision maker must choose whatever action is more likely to (causally) produce desired outcomes.
\item  His formulation falls short since he does not consider any kind of formal causal model beyond what is commonly understood by causality
\end{itemize}
\end{frame}
\subsection{Approaches and Related Work: Bandits}
\begin{frame}
\frametitle{Bandits: Non-causal}
\begin{itemize}
\item In the class of decision problems considered for this proposal it is assumed that the a rational agent must choose some available action and then receive a reward from it.
\item Problems of this nature have been modelled as \textit{bandit} problems.
\item Best-arm algorithms exist: none consider causal information.
\end{itemize}
\end{frame}
\begin{frame}
\frametitle{Bandits: Causal bandits}
\begin{itemize}
\item An action could be conceived as an \textit{intervention} over some environment, which is in fact how humans consider actions in the world
\item They use and modify causal knowledge during a sequential decision making process.
\item It is shown by \cite{lattimoreNIPS2016} that adding causal information in a fixed budget decision problem allows the decision maker to learn \textit{faster} had he not considered causal information.
\item Their work requires that the causal model is fully known to the decision maker.
\item this requirement is relaxed later by \cite{sen2017identifying} who requires only that some part of the causal model is known and allow interventions over the unknown part. 
\end{itemize}
\end{frame}

\section{Proposed Solution and Methodology}
\begin{frame}
\frametitle{Proposed Solution and Methodology}
\end{frame}
	\subsection{Objectives}
	\begin{frame}
	\frametitle{General Objectives}
	The proposed research has as a general objective:
	\begin{itemize}
 \item Understand the implications of causality for rational decision maker who faces an uncertain environment and how causal relations can be discovered and used in order to make good choices that maximize the expected utility for the decision maker.
	\end{itemize}
	\end{frame}

\subsection{Methodology}
\begin{frame}
\frametitle{Proposed Solution: Conceptual step-by-step methodology}
\begin{itemize}
\item Rational agent in a causal environment.
\item Modelling agent-environment interaction as a \textit{game}.
\item Encoding current causal knowledge as probabilsitic \textit{beliefs}
\item Using current knowledge to make a choice.
\item Updating causal information.
\end{itemize}
\end{frame}

\subsection{Validation}
\begin{frame}
\frametitle{Validation}
How to know if objective has been achieved?
\end{frame}

\begin{frame}
\frametitle{Concurrent Validation}
We will be comparing against standard Reinforcement Learning methods such as Q-Learning (\cite{watkins1992q}) since it has been shown that optimal policies achieve the objective of Max Expected Utility.
\end{frame}

\begin{frame}
\frametitle{A conjecture}
It needs to be shown that given the rewards obtained by a decision-making procedure that achieves the MEU, then my proposed algorithm must stay close to it.
\begin{teo}
Let $(X_1,X_2,...)$ the rewards obtained by a decision-making procedure which is known to converge to the max expected utility (or an optimal policy), then, if $(Y_1, Y_2,...)$ are the rewards obtained by a decision-making procedure that uses causal information, then for all $\varepsilon > 0$ there exists an $N_\varepsilon \in \mathbb{N}$ such that $ | X_t - Y_t | < \varepsilon$ for any $t > N_\varepsilon$. 
\end{teo}
\begin{proof}
To be proved.
\end{proof}
\end{frame}

\begin{frame}
In this case, if we succeed in finding an action whose reward equals that of RL algorithms, then we will have chosen the optimal action using Causal Information as proposed in the Hypothesis.
\end{frame}

\begin{frame}
\frametitle{How to}
How to
\end{frame}

\subsection{Proposed Solutuion: A guiding principle}
\begin{frame}
\frametitle{A guiding principle for Causal Decision Problems}
We propose the following principle as a normative guide for how causal information must be used in order to find a solution to a decision problem where the preferences of the decision maker are assumed to be rational. We consider only the case where the decision maker's utility is given by the outcome of a $\{0,1 \}$-binary variable. It remains to be refined and extended for a general case. 
\end{frame}

\begin{frame}
\begin{prop}
Let  $\mathcal{G}$ a causal graphical model and let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem under uncertainty whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$. Let $Y$ a variable in $\mathcal{G}$ the variable whose realizations correspond to the consequences of the decision problem once the agent has chosen an action $a \in \mathcal{A}$ and assume that $Y$ only takes values in the set $\{ 0,1\}$ where $1$ is a more desired outcome for the agent. Then, the $\textit{solution}$ for this decision problem is given by the action $a^\ast \in \mathcal{A}$ that satisfies:
\[ P(Y=1 | do(a^\ast) \geq P(Y=1 | do(a)) \textrm{ for all } a \in \mathcal{A}. \]
\end{prop}
\end{frame}

\subsection{First Step: Modelling a Causal Decision Problem as a Game}
\begin{frame}
\frametitle{Games}
\begin{itemize}
\item In this way, the first step in order to solve the stated problem is to model the interaction of the (original) decision maker and his environment as a \textbf{game} between two players: 
\begin{enumerate}
\item the (original) decision maker, 
\item and a player which will be called Nature, whose actions are to be guided by the causal model.
\end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}
\begin{itemize}
\item  Since in decision problems under uncertainty it is assumed that the \textit{state} of the environment is unknown, we will assume that player Nature has the first move and he assigns some state to any of the variable in the causal model $\mathcal{G}$.
\item After the decision maker makes his \textit{play}, Nature will \textit{respond} to the action selected according to the causal relations expressed by $\mathcal{G}$.
\item This action-response dynamic forces to consider some notion of order between the players' moves, and because of this reason we must use the \textit{extensive-form} games
\end{itemize}
\end{frame}
